Efficient management and analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr

### 2017-09-19 cabios-template/draft.tex
update last results

%The Celiac dataset is relatively small as compared to modern genetic cohorts. 
To illustrate the scalability of the two R packages, we performed a GWAS analysis on 500K individuals and 1M SNPs. The GWAS analysis completed in less than 5 hours using the aforementioned desktop computer. The GWAS analysis was composed of three main steps. First, we removed SNPs in long-range LD regions and used SNP clumping, leaving 93,083 SNPs. Then, the 10 first PCs were computed on the 500K individuals and these remaining SNPs. Finally, we performed a linear association test in the complete 500K dataset for each of the 1M SNPs, using the 10 first PCs as covariables.

@@

On our desktop computer, we compared the computation times of FastPCA, FlashPCA2 to the similar function big\_randomSVD implemented in bigstatsr. For each comparison, we used the 93,083 SNPs which were remaining after pruning and we computed 10 PCs. We used the datasets of growing size simulated from the Celiac dataset. Overall, our function big\_randomSVD showed to be twice as fast as FastPCA and FlashPCA2 and almost 10 times as fast when using parallelism (an option not currently possible with either FastPCA or FlashPCA2) with 6 cores (Figure~\ref{fig:bench-pca}). We also compared results in terms of precision by comparing squared correlation between approximated PCs and ``true'' PCs provided by an exact singular value decomposition obtained with SmartPCA. FastPCA, FlashPCA2 and bigstatsr infer the true first 6 PCs but the squared correlation between true PCs and approximated ones decreases for larger PCs when using FastPCA (Fast mode of EIGENSOFT) whereas it remains larger than 0.999 when using FlashPCA2 or bigstatsr (Figure~\ref{fig:prec-pca}).

@@

For the imputation method based on XGBoost, we compared  the imputation accuracy and computation times with Beagle on the POPRES dataset. The histogram of the minor allele frequencies (MAFs) of this dataset is provided in figure~\ref{fig:maf} and there is no missing value. 
We used a Beta-binomial distribution to simulate the number of missing values by SNP and then randomly introduced missing values according to these numbers, resulting in approximately 3\% of missing values overall (Figure~\ref{fig:NA}). Imputation was compared between function snp\_fastImpute of package bigsnpr and Beagle 4.1 (version of January 21, 2017). Overall, snp\_fastImpute made 4.7\% of imputation errors whereas Beagle made only 3.1\% of errors but it took Beagle 14.6 hours to complete while our method only took 42 minutes (20 times less). 
We also show that the estimation of the number of imputation errors is accurate (Figure~\ref{fig:error-impute}).
For the Celiac dataset in which there was already missing values, in order to further compare computation times, we report that snp\_fastImpute took less than 6 hours to complete for the whole genome whereas Beagle didn't finish imputing chromosome 1 in 48 hours.

>>>>

%The Celiac dataset is relatively small as compared to modern genetic cohorts. 
To illustrate the scalability of the two R packages, we performed a GWAS analysis on 500K individuals and 1M SNPs. The GWAS analysis completed in approximately 11 hours using the aforementioned desktop computer. The GWAS analysis was composed of four main steps. 
First we read from PLINK files in our format "bigSNP" in 1 hour.
Then, we removed SNPs in long-range LD regions and used SNP clumping, leaving 93,083 SNPs in 5.4h. Then, the 10 first PCs were computed on the 500K individuals and these remaining SNPs in 1.8h. Finally, we performed a linear association test on the complete 500K dataset for each of the 1M SNPs, using the 10 first PCs as covariables in 2.9h.

@@

On our desktop computer, we compared the computation times of FastPCA, FlashPCA2 to the similar function big\_randomSVD implemented in bigstatsr. For each comparison, we used the 93,083 SNPs which were remaining after pruning and we computed 10 PCs. We used the datasets of growing size simulated from the Celiac dataset. Overall, our function big\_randomSVD showed to be almost twice as fast as FastPCA and FlashPCA2 and 8 times as fast when using parallelism (an option not currently possible with either FastPCA or FlashPCA2) with 6 cores (Figure~\ref{fig:bench-pca}). We also compared results in terms of precision by comparing squared correlation between approximated PCs and ``true'' PCs provided by an exact singular value decomposition obtained with SmartPCA. FastPCA, FlashPCA2 and bigstatsr infer the true first 6 PCs but the squared correlation between true PCs and approximated ones decreases for larger PCs when using FastPCA (Fast mode of EIGENSOFT) whereas it remains larger than 0.999 when using FlashPCA2 or bigstatsr (Figure~\ref{fig:prec-pca}).

@@

For the imputation method based on XGBoost, we compared  the imputation accuracy and computation times with Beagle on the POPRES dataset. The histogram of the minor allele frequencies (MAFs) of this dataset is provided in figure~\ref{fig:maf} and there is no missing value. 
We used a Beta-binomial distribution to simulate the number of missing values by SNP and then randomly introduced missing values according to these numbers, resulting in approximately 3\% of missing values overall (Figure~\ref{fig:NA}). Imputation was compared between function snp\_fastImpute of package bigsnpr and Beagle 4.1 (version of January 21, 2017). Overall, snp\_fastImpute made 4.7\% of imputation errors whereas Beagle made only 3.1\% of errors but it took Beagle 14.6 hours to complete while our method only took 42 minutes (20 times less). 
We also show that the estimation of the number of imputation errors is accurate (Figure~\ref{fig:error-impute}).
For the Celiac dataset in which there was already missing values, in order to further compare computation times, we report that snp\_fastImpute took less than 10 hours to complete for the whole genome whereas Beagle didn't finish imputing chromosome 1 in 48 hours.
